---
title: "Payroll and Team's Success in Major League Baseball (MLB)"
subtitle: "Exploratory Data Analysis (EDA) of MLB Payroll Data"
author: "Calvin Brauer, Jihan Lee"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: simplex
    toc: true
    number_sections: true
    toc_float:
      collapsed: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

# Introduction
In our analysis, we are examining data 5 years of data from Major League Baseball (MLB), specifically focusing on the correlation between team expenditures on player salaries and their on-field performance. One distinctive feature of the MLB, compared to other American professional sports leagues, is the absence of a strict salary cap. This absence raises an intriguing question: Is it feasible for teams to achieve success merely by allocating substantial financial resources to secure top-tier players? Our analysis aims to ascertain the extent to which financial investment in player payroll translates to on-field victories.

# Data
Our project delves into datasets from spotrac, sourced from their Major League Baseball (MLB) payroll page (https://www.spotrac.com/mlb/payroll/). These datasets encompasses payroll details for all 30 MLB teams across five seasons: 2018-19, 2021-23. Notably, we excluded the 2020 season due to its shortened season by the COVID-19 pandemic.

Our analysis benefits from the inclusion of the recently concluded 2023 season. It is important to recognize that these five years of data do not necessarily represent trends in preceding seasons or predict future patterns.

An ethical aspect to consider is the nature of the dataset's acquisition: it was scraped from the web. The original purpose of the data was to promote transparency within the league. Our use of it is solely for academic exploration and insights, and we urge caution and ethical consideration in any potential commercial use.

The dataset, while compact, is well-structured and requires minimal preprocessing. We anticipate minor data cleaning tasks like renaming certain columns or data points for clearer analysis. Our exploratory data analysis aims to identify patterns, anomalies, and insights in the MLB payroll figures for these years.

A significant addition to our analysis is the incorporation of five years of Consumer Price Index (CPI) data (https://fred.stlouisfed.org/series/CPIAUCSL). This inclusion enables us to make consistent comparisons between different years' data by adjusting for inflation, providing a more accurate understanding of payroll changes over time.

To initiate our comprehensive analysis of the MLB Team Payroll, we should start by importing the datasets.

```{r import datasets, warning=FALSE, message=FALSE}
library('readxl')
library('tidyverse')
library('knitr')

payrollPosition18 <- read_excel("../data/mlb_payroll_tracker_positional_breakdown_2018.xlsx")
payrollPosition19 <- read_excel("../data/mlb_payroll_tracker_positional_breakdown_2019.xlsx")
payrollPosition21 <- read_excel("../data/mlb_payroll_tracker_positional_breakdown_2021.xlsx")
payrollPosition22 <- read_excel("../data/mlb_payroll_tracker_positional_breakdown_2022.xlsx")
payrollPosition23 <- read_excel("../data/mlb_payroll_tracker_positional_breakdown_2023.xlsx")

payrollRoster18 <- read_excel("../data/mlb_payroll_tracker_roster_breakdown_2018.xlsx")
payrollRoster19 <- read_excel("../data/mlb_payroll_tracker_roster_breakdown_2019.xlsx")
payrollRoster21 <- read_excel("../data/mlb_payroll_tracker_roster_breakdown_2021.xlsx")
payrollRoster22 <- read_excel("../data/mlb_payroll_tracker_roster_breakdown_2022.xlsx")
payrollRoster23 <- read_excel("../data/mlb_payroll_tracker_roster_breakdown_2023.xlsx")
```

The data for each season is derived from two distinct but similar datasets, each structured around the 30 MLB teams. The `payrollPosition` dataset breaks down payroll according to various defensive position groupings, whereas the `payrollRoster` dataset categorizes payroll based on different roster classifications.

## Pre-processing
In our data pre-processing, we have identified a specific entry labeled `LEAGUE AVERAGE` within the `payrollRoster` dataset. While this entry provides a valuable benchmark by offering an average across the league, its inclusion may introduce ambiguity, especially when conducting team-specific analyses. To ensure the clarity of our data and to avoid potential misinterpretations or skewness in subsequent analyses, we are going to eliminate this entry. By doing so, we aim to maintain a dataset that solely represents individual team figures.

```{r filter, warning=FALSE, message=FALSE}
filterPayrollRoster <- function(dataFrame) {
  filteredDataFrame <- dataFrame %>%
    filter(RANK != 'LEAGUE AVERAGE')
  return(filteredDataFrame)
}

rosterNames <- c("payrollRoster18", "payrollRoster19", "payrollRoster21", "payrollRoster22", "payrollRoster23")

for (name in rosterNames) {
  assign(name, filterPayrollRoster(get(name)))
}
```

In our data cleaning process, we have streamlined our approach by utilizing lists for different years of datasets and looping through a series of following steps to reduce code repetition. Initially, to address inconsistencies within the datasets, we will transform the `RANK` variable into an integer format. Also, we will set `INJURED RESERVE` and `DH $` variables into numeric variables for each dataset since they are originally imported as character types. Following this, we are going to employ the `full_join()` method to merge each year's two datasets (payroll positions and rosters) into a single cohesive dataset. This merging is an essential step to consolidate the data for comprehensive analysis.

Further enhancing our datasets' utility, we will add a `year` variable to each year's dataset. This addition allows for easy identification and comparison in later stages of analysis, particularly when assessing trends or changes over the years. Finally, before merging the datasets from different years into one comprehensive dataset, we will take the step of renaming variables for consistency, which ensures that our final, combined dataset, achieved through the `rbind()` function, is well-organized and uniform.

```{r clean and merge datasets, warning=FALSE}
payrollRosterList <- list(payrollRoster18, payrollRoster19, payrollRoster21, payrollRoster22, payrollRoster23)
payrollPositionList <- list(payrollPosition18, payrollPosition19, payrollPosition21, payrollPosition22, payrollPosition23)

payrollDataNames <- c("payrollData18", "payrollData19", "payrollData21", "payrollData22", "payrollData23")

for (i in 1:length(payrollRosterList)) {
  payrollRosterList[[i]]$RANK <- as.integer(payrollRosterList[[i]]$RANK)

  payrollData <- full_join(payrollPositionList[[i]], payrollRosterList[[i]], by = c('TEAM', 'RANK', 'ROSTER'))
  
  assign(payrollDataNames[i], payrollData)
}

payrollData18$year <- 2018
payrollData19$year <- 2019
payrollData21$year <- 2021
payrollData22$year <- 2022
payrollData23$year <- 2023

payrollData18 <- payrollData18 %>% rename('total'='2018 TOTAL PAYROLL')
payrollData19 <- payrollData19 %>% rename('total'='2019 TOTAL PAYROLL')
payrollData21 <- payrollData21 %>% rename('total'='2021 TOTAL PAYROLL')
payrollData22 <- payrollData22 %>% rename('total'='2022 TOTAL PAYROLL')
payrollData23 <- payrollData23 %>% rename('total'='2023 TOTAL PAYROLL')

payrollData18 <- payrollData18 %>% rename('normalized'='25-MAN PAYROLL')
payrollData19 <- payrollData19 %>% rename('normalized'='25-MAN PAYROLL')
payrollData21 <- payrollData21 %>% rename('normalized'='26-MAN PAYROLL')
payrollData22 <- payrollData22 %>% rename('normalized'='26-MAN PAYROLL')
payrollData23 <- payrollData23 %>% rename('normalized'='26-MAN PAYROLL')

payrollData <- rbind(payrollData18, payrollData19, payrollData21, payrollData22, payrollData23)
```

We will now clean up the data further by eliminating variables that are named differently but are essentially duplicates of each other.

```{r select variables}
payrollData <- payrollData %>%
  select('year', 'RANK', 'TEAM', 'WIN %', 'ROSTER', 'CATCHER $',
         'INFIELD $', 'OUTFIELD $', 'DH $', 'PITCHING $',
         'total', 'normalized', 'INJURED RESERVE', 'RETAINED', 'BURIED')
```

For the purpose of ensuring consistency and facilitating subsequent analysis, we will standardize the naming of variables. To achieve this, we have adopted the camelCase naming convention.

```{r rename variables}
payrollData <- payrollData %>%
  rename(
    rank = 'RANK',
    team = 'TEAM',
    winPercentage = 'WIN %',
    roster = 'ROSTER',
    catcher = 'CATCHER $', 
    infield = 'INFIELD $',
    outfield = 'OUTFIELD $',
    dh = 'DH $',
    pitching = 'PITCHING $',
    injuredReserve = 'INJURED RESERVE',
    retained = 'RETAINED',
    buried = 'BURIED'
  )
```

During our data cleaning process, we have encountered a unique situation with the team name change of the Cleveland Indians to the Cleveland Guardians in 2021. This change was a response to growing concerns about the perpetuation of Native American stereotypes. Recognizing the importance of maintaining consistency in our analysis, especially when dealing with data spanning both before and after this name change, we have decided to update the team name across all datasets. To achieve this, we are going to utilize the `mutate()` function in R. This function will allow us to seamlessly modify the team name in our datasets, replacing "Cleveland Indians" with "Cleveland Guardians" for all relevant entries, regardless of the year.

```{r rename values}
payrollData <- payrollData %>%
  mutate(team = ifelse(team == "Cleveland Indians", "Cleveland Guardians", team))
```

To ensure data integrity and consistency, it is necessary to verify that each variable is of the appropriate class. The `sapply()` method provides a concise display for this purpose.

```{r check classes of varible}
sapply(payrollData, class)
```

Upon examining the output, it is evident that most variables align with their expected classes. However, there are exceptions with the `dh` and `injuredReserve` variables, which are currently of class "character" but should ideally be "numeric". We will proceed to fix this discrepancy by adjusting the data class of these variables.

```{r change variable class, warning=FALSE}
payrollData$dh <- as.numeric(payrollData$dh)
payrollData$injuredReserve <- as.numeric(payrollData$injuredReserve)
```

In our data cleaning process, a crucial step is adjusting for inflation to ensure that our analysis is accurate and meaningful over time. We will begin this process by importing the Consumer Price Index (CPI) dataset from the Federal Reserve Economic Data (FRED).

To make our analysis more intuitive and straightforward, we will first rename the columns of the CPI dataset and change the data type of the `index` column to double. Given that our data includes two data points per year, we will filter the CPI data to retain only the index value from the beginning of each year. Afterwards, we are going to alter the date format to display only the year to simplify the display.

Next, we will calculate inflation factors for each year by dividing `baseIndex`, which we set as the 2023 index with a value of 100, by the index of each respective year. We then compile these factors into a vector. Finally, using the `mutate()` function, we adjust all the columns in our dataset that represent dollar values for each year, applying the inflation factors accordingly.

```{r inflation adjustment, message=FALSE, warning=FALSE}
cpi <- read_csv("../data/CPIAUCSL.csv")

cpi <- cpi %>% rename(date = 'DATE', index = 'CPIAUCSL_NBD20230101')
cpi$index <- as.double(cpi$index)

cpi <- cpi %>%
  filter(format(as.Date(date), "%m-%d") == "01-01") %>%
  mutate(date = format(as.Date(date), "%Y"))

baseIndex <- cpi[cpi$date==2023, 'index']

infl1823 <- baseIndex / cpi[cpi$date==2018, 'index']
infl1923 <- baseIndex / cpi[cpi$date==2019, 'index']
infl2123 <- baseIndex / cpi[cpi$date==2021, 'index']
infl2223 <- baseIndex / cpi[cpi$date==2022, 'index']
infl2323 <- baseIndex / cpi[cpi$date==2023, 'index']

inflFactors <- c('2018' = infl1823, '2019' = infl1923, '2021' = infl2123, '2022' = infl2223, '2023' = infl2323)
inflFactors <- unlist(inflFactors)

payrollData <- payrollData %>%
  mutate(
    across(c(catcher, infield, outfield, dh, pitching, total, normalized, injuredReserve, retained, buried),
           ~ .x * sapply(year, function(y) inflFactors[paste0(y, ".index")]))
  )
```

## Overview of Varibales
As we have pre-processed the data, we can now delve deeper into understanding the individual variables. While a comprehensive exploration of every variable is beyond the scope of this initial overview, we will emphasize those variables with most significance:

* `year`: Indicates the year the season happened.
* `rank`: Represents the team's standing, numerically ranging from 1 to 30, based on the magnitude of their total payroll.
* `team`: Denotes the official name of the MLB team.
* `winPercentage`: Shows the proportion of games won by the team during the regular season.
* `roster`: Quantifies the total number of players within the team's roster.
* `total`: Reflects the total amount of salary spent by the team towards their roster for the year.
* `normalized`: Signifies the amount of salary associated with the team's 25-man or 26-man roster, ensuring a normalized payroll comparison irrespective of the total roster size.

## Central Tendency Measures and Descriptive Statistics

```{r central tendency measures}
meanTotal <- format(round(mean(payrollData$total), 2), nsmall = 2)
medianTotal <- format(round(median(payrollData$total), 2), nsmall = 0)
sdTotal <- format(round(sd(payrollData$total), 2), nsmall = 2)
```

The payroll landscape of the MLB teams over the span of 5 seasons can be summarized through a few key descriptive statistics:

* Average (Mean) Total Payroll: The arithmetic mean provides a central value, giving us a general idea of the typical team payroll. For our dataset, the average payroll stands at **$`r meanTotal`**. This figure indicates the central tendency of the payroll data and offers a reference point against which individual team payrolls can be compared.

* Median Total Payroll: The median, being the middle value when all numbers are sorted in ascending or descending order, is less sensitive to extreme values or outliers than the mean. The median payroll for our dataset is **$`r medianTotal`**, suggesting that half of the MLB teams spent more than this amount and half spent less. The comparison between the mean and median can also provide insights into the skewness of the distribution. Given that the average total payroll exceeds the median, this suggests a slight right-skew in the data. This skewness implies the presence of certain teams that have significantly higher payrolls, which pull the overall average upwards. These outliers indicate a disparity where a select few teams allocate considerably more funds than the majority.

* Standard Deviation of Total Payroll: The standard deviation quantifies the extent to which individual payrolls deviate from the mean. A higher standard deviation indicates a wider dispersion of values around the mean. In our dataset, the standard deviation is **$`r sdTotal`**. This value gives us an indication of the variability or spread of the team payrolls around the average.

Now we will delve into the summary statistics for the variables contained within this dataset.

```{r summary statistics}
summary(payrollData)
```

By understanding these metrics, we can gain a deeper evaluation of the financial dynamics and variations across different MLB teams over the 5 seasons.

## Distributions of Data {.tabset}
Visualizing the various distributions within our dataset will offer distinct insights, guiding our analytical approach. Before delving into this, we must first import the necessary packages to facilitate such visualizations.

```{r import ggplot2, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggthemes)
library(gridExtra)
theme_set(theme_clean())
```

### Line Plot
In this part, we have chosen to utilize a line plot as it is particularly effective in illustrating trends over time. We will create a line plot for examining the trends of the median total and normalized amounts in the payroll data. We choose to use the median as the central measure in our line plot because, as highlighted in our descriptive statistics, the average (or mean) demonstrated a degree of skewness.

```{r line plot}
aggData <- payrollData %>%
  group_by(year) %>%
  summarize(total = median(total), normalized = median(normalized))

ggplot(aggData, aes(x=year)) + 
    geom_line(aes(y=total, color="Total Payroll"), size=2) + 
    geom_line(aes(y=normalized, color="26-Man Payroll"), size=2) +
    scale_color_manual(name="Payrolls",values=c("#002D72", "#D50032"))+
    labs(title="Median Total vs. 26-Man Payroll Over Years", x="Year", y="Payroll") +
    scale_y_continuous(labels = scales::dollar_format(scale = 1e-6, prefix = "$", suffix = "M"), breaks = seq(0, 200000000, 25000000))
```

The blue line in the plot depicts the normalized or 26-man payroll, while the red line illustrates the total payroll. We observed that the overall trends for both the normalized and total payroll over time were similar. It is important to note that the 2020 data point in our plot is an interpolation. We had previously decided not to include the data for the year 2020 in our analysis. As such, the 2020 point on the graph has been estimated based on surrounding data to maintain continuity in the visual trend.

Surprisingly, despite expectations to the contrary, both variables exhibited a decreasing trend over time after adjusted to be real terms. The downward trend in both the normalized and total payroll opens up intriguing avenues for further investigation into the underlying factors driving this unexpected pattern.

### Histogram
A histogram is a helpful tool for analyzing the distribution of numeric variables, providing a clear visual representation of data spread and frequency. Here, we will use a histogram to examine the distribution of total payrolls.

```{r histogram}
ggplot(payrollData, aes(x=total)) + 
  geom_histogram(bins=20, fill="#002D72", color="#D50032") +
  labs(title="Distribution of Total Payrolls", x="Real Total Payroll", y="Frequency") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-6, prefix = "$", suffix = "M"), breaks = seq(0, 350000000, 50000000))
```

Overall, the histogram depicting the distribution of total payrolls reveals a slight right-skewedness, centering around an average of $150 million in real terms. This right skewness in the distribution suggests that a few teams have invested significantly more in their payroll compared to the majority, indicating a disparity in payroll expenditures among the teams. This pattern highlights the presence of outliers on the higher end of the payroll spectrum.

### Boxplots
To delve deeper into the nuances of our data, we will employ boxplots to show the distributions of various variables. While the line plot we used previously effectively highlighted the overall trend of the median total payroll over the years, boxplots offer a different perspective. By plotting the year on the x-axis and the total payroll on the y-axis, boxplots will enable us to examine the changes in distribution over time. 

```{r boxplot, warning=FALSE, message=FALSE}
ggplot(payrollData, aes(x=factor(year), y=total)) +
  geom_boxplot(fill="#002D72", color="#D50032") +
  labs(title="Total Payroll Over Years", x="Year", y="Total Payroll") +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-6, prefix = "$", suffix = "M"), breaks = seq(0, 350000000, 50000000))
```

As observed in the line plot, there is a discernible slight decreasing trend in our data. Notably, the boxplot analysis reveals that the data's spread varies, with the year 2023 exhibiting the highest scope of dispersion.

Now, we will shift our focus to analyzing the payroll breakdown by different positions. To facilitate this analysis, we will utilize the `melt()` function from the `reshape2` package, which will transform our dataset into a long format. In the resulting visualization, the x-axis will represent the total payroll in each position, while the y-axis will display the different positions. This approach allows us to compare payrolls across various positions directly, providing clear insights into how resources are distributed among different roles within the team.

```{r boxplots by position, warning=FALSE, message=FALSE}
library(reshape2)
payrollLong <- melt(payrollData, id.vars = c("year", "team"), 
                    measure.vars = c("catcher", "pitching", "infield", "outfield", "dh"),
                    variable.name = "position", value.name = "payroll")
payrollLong$payroll <- as.numeric(payrollLong$payroll)

payrollLong <- payrollLong %>% 
  group_by(year, team, position) %>%
  summarize(payroll = sum(payroll))

ggplot(payrollLong, aes(x=payroll, y=position)) +
  geom_boxplot(fill="#002D72", color="#D50032") +
  labs(title="Payroll Breakdown by Position", x="Total Payroll", y="Position") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-6, prefix = "$", suffix = "M"), breaks = seq(0, 350000000, 50000000)) +
  scale_y_discrete(labels = c("dh" = "DH", 
                              "outfield" = "Outfielder",
                              "infield" = "Infielder",
                              "pitching" = "Pitcher",
                              "catcher" = "Catcher"))
```
In our visualization of the payroll breakdown by different positions, the overall trend mirrors the right skewness observed in the total payroll distribution. This skewness is indicative of a few significantly high values pulling the average upwards. In terms of specific positions, it is evident that pitchers generally command the highest salaries, whereas catchers tend to earn the least. This disparity in pay across positions is a notable aspect of our findings. Additionally, the analysis reveals a number of outliers in the salary distributions for outfielders, pitchers, and catchers. These outliers are particularly interesting, as they suggest that some teams pay significantly more certain positions, which might depend on each team's different resource allocation strategy.

These insights provide a nuanced understanding of team expenditures and their distribution strategies.

### Scatter Plot
Scatter plots are helpful in illustrating the overarching trend between two continuous variables. In our context, we are particularly interested in examining the relationship between payroll amounts and win percentages. Consequently, we will create a scatter plot for these variables, accompanies by a line of best fit, to discern any significant correlation or trend between them.

```{r scatter plot, warning=FALSE, message=FALSE}
ggplot(payrollData, aes(x=total, y=winPercentage)) +
  geom_point(color="#002D72") +
  geom_smooth(method=lm, color="#D50032") +
  labs(x="Total Payroll", y="Win %", title="Total Payroll vs Win Percentage") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-6, prefix = "$", suffix = "M"), breaks = seq(0, 350000000, 50000000))
```

From the given graph, it is evident that there is a positive association between the Total Payroll and Win Percentages. This suggests that teams with higher expenditures on their players generally exhibit superior on-field results. However, the visual representation alone does not quantify the strength of this relationship. To precisely gauge this association, we will proceed with a correlation analysis on these two variables.

```{r correlation analysis}
library(broom)
corrtest <- cor.test(payrollData$total, payrollData$winPercentage, method="pearson")
tidy(corrtest)
```

From the above result, the correlation coefficient is 0.421, which indicates a moderate positive relationship between the two variables. In other words, as total payroll increases, the win percentage tends to increase as well. Additionally, given that the p-value of this estimate is 8.309767e-08, which is substantially below conventional significance levels (0.5, 0.1, 0.05, 0.01), we can confidently assert that this correlation is statistically significant. This implies that the observed correlation coefficient is not a mere coincidence but rather an indicative measure of an underlying association between the variables.

However, it is crucial to note that correlation does not imply causation. While there is a strong correlation between payroll and team performance, it does not necessarily mean that higher payrolls directly cause better performances. To draw causal inferences, we would need to consider additional assumptions and employ different analytical methodologies.

# Additional Topical Analysis
Building on the insights from our initial exploratory data analysis, we will now undertake a more detailed examination of the topics of our interest. In this section, we have chosen to exclusively utilize data from the year 2022. This decision was made for clarity and convenience, allowing us to provide a more focused and understandable examination of the payroll structure within that specific year. However, it is important to note that this is a preliminary step, and we plan to update our analysis in the future to incorporate more data from additional years. The incorporation of data from multiple years will significantly enhance the depth and accuracy of our analysis.

## Above Average Pay and Wins
Firstly, we will examine teams that distinguished themselves with both above-average payrolls and superior winning percentages.

```{r}
payrollData[payrollData$year == 2022, ] %>%
  filter(total > mean(total)) %>%
  filter(winPercentage > mean(winPercentage))
```

Seven teams stood out with both above-average payrolls and winning percentages. Notably, among these, the top three highest-paying teams all showed superior winning percentages.

## Below Average Pay and Wins
Similarly, we will look at teams that had both below-average payrolls and wins.

```{r}
payrollData[payrollData$year == 2022, ] %>%
  filter(total < mean(total)) %>%
  filter(winPercentage < mean(winPercentage))
```

Interestingly, there were nine teams with both below-average payrolls and winning percentages. Contrasting with the teams that excelled in both payroll and victories, these teams displayed a broader range in their standings. For example, despite the Baltimore Orioles having the lowest payroll in the MLB, they secured victories in approximately 52% of their season's matches, surpassing the average win rate. However, it is evident that a majority of the teams falling below the average in performance also operated on modest budgets. For instance, the Washington Nationals, with the 10th lowest payroll, found themselves at the bottom of the league, winning just over a third of their matches.

## Teams That Were "Unlucky"
In sports, injuries introduce an element of unpredictability. Teams can occasionally face the misfortune of their best-performing players, often the highest earners, having long-term injuries, sidelining them for significant portions of the season. In such instances, these players are placed to the injured reserve list, freeing up a roster spot for another player to replace their roster position. Our analysis will compare the proportion of a team's payroll allocated to players on the injured reserve against the team's overall winning percentage.

```{r warning=FALSE, message=FALSE}
payrollData[payrollData$year == 2022, ] %>%
  mutate(pctIR = injuredReserve / total) %>%
  ggplot(aes(x=pctIR, y = winPercentage)) +
  geom_point(color="#002D72") +
  geom_smooth(method=lm, color="#D50032") +
  labs(x="% of Payroll Towards Injured Reserve", y="Win %", title="Injured Reserve Payroll vs Win Percentage")
```

There appears to be a slight negative corrleation between injured reserve utilization and win percentage. The downward-sloping trend line suggests that having a higher proportion of a team's payroll going towards players on the injured reserve could be linked to poorer performance on the field.

## What Defines Success?
In our analysis thus far, we have primarily defined success in terms of winning percentage. Yet, is this the best metric for gauging success? While having the most wins during the regular season is great, it is championship wins that record a team's legacy in history. Notably, the Houston Astros won the 2022 MLB World Series, beating the Philadelphia Phillies. The Phillies, despite not securing the regular season title, became National League champions and advanced to the World Series. Furthermore, teams that dominated the regular season, such as the Los Angeles Dodgers and Atlanta Braves, did not even advance to the Championship Series.

```{r}
payrollData[payrollData$year == 2022, ] %>%
  filter(team == 'Houston Astros' | team == 'Philadelphia Phillies')
```

While the given table might not seem the most interesting, it offers some valuable insights. Specifically, having the richest team and investing a lot of money on players does not necessarily guarantee the World Series title. Neither of the finalists featured in the season's top three highest payrolls. The champion Astros, ranking eighth, were essentially mid-tier, given that there are 30 teams in the league.

# Discussion
For this exploratory data analysis (EDA) project, we acknowledge the need to extend our study by incorporating additional years of data. This expansion will enable us to analyze longer-term trends and gain a more comprehensive understanding of the dynamics at play. Given that our dataset constitutes panel data, it opens up opportunities for more sophisticated time series analyses. Methods such as Granger causality tests could be instrumental in uncovering the ability to forecast between variables within the data. Furthermore, employing causal inference methodologies like fixed effects or random effects models could provide deeper insights into the underlying mechanisms.

However, it is crucial to recognize that drawing robust causal conclusions from such analyses demands a more rigorous analytical framework. This entails the incorporation of control variables and an acute awareness of potential confounders that could influence the outcomes. By carefully considering these factors, we can significantly enhance the validity and reliability of our findings.
